---
title: "psoConstructoin_genomicInfoFormating"
output: html_document
date: "2025-03-18"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries
```{r}
library(tidyverse)
library(taxonomizr)
library(vroom)
library(phyloseq)
library(microViz)


options(scipen=999) #For consistent p-value evaluation 
`%notin%` <- function(x,y) !(x %in% y)
```

## Taxonomic Consensus: 

VC3
```{r}
vc3 <- vroom(file = "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/vc3_final_assignments.csv")

vc3 <- vc3 %>%
  filter(grepl("contig", GenomeName) == TRUE |
           grepl("edge", GenomeName) == TRUE | 
           grepl("virus_comp", GenomeName) == TRUE) %>%
  select(GenomeName, Proteins, `Size (Kb)`, `realm (prediction)`,
         `phylum (prediction)`, `class (prediction)`, `order (prediction)`,
         `family (prediction)`, `subfamily (prediction)`, `genus (prediction)`, network) 

colnames(vc3) <- c("contigID", "vc3_num_proteins", "length", "vc3_realm", "vc3_phylum", "vc3_class", "vc3_order", "vc3_family", "vc3_subfamily", "vc3_genus", "vc3_network")

final_vc3 <- vc3 %>%
  filter(vc3_realm != "No Realm") %>%
  filter(vc3_realm != "No prediction") %>%
  mutate(across(everything(), ~ifelse(grepl("novel", .) == TRUE, NA, .))) %>%
  mutate(vc3_superkingdom = "Viruses",
         vc3_phylum = ifelse(grepl("Phixviricota", vc3_phylum) == TRUE,
                             "Phixviricota", vc3_phylum),
         vc3_class = ifelse(grepl("Malgrandaviricetes", vc3_class) == TRUE,
                             "Malgrandaviricetes", vc3_class)) %>%
  select(contigID, vc3_superkingdom, vc3_phylum, vc3_class, vc3_order, vc3_family, vc3_subfamily, vc3_genus)

final_vc3 %>%
  group_by(vc3_phylum) %>% count(vc3_phylum)
```

MMSeqs
```{r}
mmseqs_LCA <- vroom(file = "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/mmseqs_vOTU_repSeqs_lca.tsv", col_names = FALSE)

mmseqs_topHit <- vroom(file = "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/mmseqs_vOTU_repSeqs_tophit_aln", col_names = FALSE)
colnames(mmseqs_topHit) <- c("query", "target", "evalue", "pident", "fident", "nident", "mismatch", "qcov", "tcov", "qstart", "qend", "qlen", "tstart", "tend", "tlen", "alnlen", "bits", "qheader", "theader", "taxid", "taxname", "taxlineage")

mmseqs_best_hits_by_bitscore <- mmseqs_topHit %>%
  filter(grepl("d_Viruses", taxlineage, ignore.case = TRUE)) %>%
  filter(pident >= 70) %>%
  group_by(query) %>%
  slice(which.max(bits)) %>%
  ungroup() %>%
  select(query, taxid)

# linage <- as.data.frame(getTaxonomy(mmseqs_best_hits_by_bitscore$taxid, "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/miscellaneous/accessionTaxa.sql"))
linage <- as.data.frame(getTaxonomy(mmseqs_best_hits_by_bitscore$taxid, "~/Documents/accessionTaxa.sql"))

mmseqs_best_hits_by_bitscore$blast_superkingdom <- linage$superkingdom
mmseqs_best_hits_by_bitscore$blast_phylum  <- linage$phylum
mmseqs_best_hits_by_bitscore$blast_class  <- linage$class
mmseqs_best_hits_by_bitscore$blast_order  <- linage$order
mmseqs_best_hits_by_bitscore$blast_family  <- linage$family
mmseqs_best_hits_by_bitscore$blast_genus  <- linage$genus
mmseqs_best_hits_by_bitscore$blast_species  <- linage$species

# Get lineage for non-'root' entries
non_root_entries <- mmseqs_LCA %>%
  filter(X4 != "root") %>%
  select(X1, taxid = X2)

# linage <- as.data.frame(getTaxonomy(non_root_entries$taxid, "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/miscellaneous/accessionTaxa.sql"))
linage <- as.data.frame(getTaxonomy(non_root_entries$taxid, "~/Documents/accessionTaxa.sql"))


non_root_entries$blast_superkingdom <- linage$superkingdom
non_root_entries$blast_phylum  <- linage$phylum
non_root_entries$blast_class  <- linage$class
non_root_entries$blast_order  <- linage$order
non_root_entries$blast_family  <- linage$family
non_root_entries$blast_genus  <- linage$genus
non_root_entries$blast_species  <- linage$species

# Checking hits with lineage in NA: taxonomic discrepancy between NCBI and my version of accessionTax
non_root_entries_NA <- non_root_entries %>%
  filter(is.na(blast_superkingdom)) %>%
  distinct(taxid)

print(non_root_entries_NA)

# Read the detailed taxonomy information from the saved file
taxonomy_details <- read.table("~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/miscellaneous/Updated_Taxonomy_Table.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Join the two dataframes by 'taxid' from non_root_entries and 'Taxonomy ID' from taxonomy_details
non_root_entries_updated <- non_root_entries %>%
  left_join(taxonomy_details, by = c("taxid" = "Taxonomy.ID")) %>%
  mutate(
    blast_superkingdom = coalesce(blast_superkingdom, Superkingdom),
    blast_phylum = coalesce(blast_phylum, Phylum),
    blast_class = coalesce(blast_class, Class),
    blast_order = coalesce(blast_order, Order),
    blast_family = coalesce(blast_family, Family),
    blast_genus = coalesce(blast_genus, Genus),
    blast_species = coalesce(blast_species, Species)
  ) %>%
  select(X1, taxid, blast_superkingdom, blast_phylum, blast_class, 
         blast_order, blast_family, blast_genus, blast_species)

# Handle 'root' entries using best viral hit
root_entries <- mmseqs_LCA %>%
  filter(X4 == "root") %>%
  select(X1) %>%
  left_join(mmseqs_best_hits_by_bitscore, by = c("X1" = "query"))

final_mmseqs <- bind_rows(non_root_entries_updated, root_entries) %>%
  select(-taxid)
colnames(final_mmseqs)[1] <- "contigID"

# Adding superkingdom to the unclassified 
final_mmseqs <- final_mmseqs %>%
  mutate(blast_superkingdom = ifelse(is.na(blast_superkingdom), "Viruses", blast_superkingdom))
```

PhaGCN
```{r}
# Load PhaGCN and formatting 
phagcn <- vroom::vroom("~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/phagcn_prediction.tsv", delim = "\t", col_names = TRUE)

phagcn <- phagcn %>%
  filter(PhaGCNScore >= 0.6)

# Define a function to parse and extract taxonomy levels
extract_taxonomy <- function(tax_string) {
  parts <- str_split(tax_string, ";", simplify = TRUE)
  levels <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")
  taxonomy <- setNames(vector("list", length(levels)), levels)
  taxonomy[] <- lapply(taxonomy, function(x) NA)
  
  for (part in parts) {
    split_part <- str_split(part, ":", simplify = TRUE)
    if (length(split_part) >= 2) {
      key <- tolower(split_part[1])
      value <- split_part[2]
      if (key %in% names(taxonomy)) {
        taxonomy[[key]] <- value
      }
    }
  }
  
  return(taxonomy)
}

final_phagcn <- phagcn %>%
  mutate(
    TaxonomyData = lapply(Lineage, extract_taxonomy),
    phagcn_superkingdom = sapply(TaxonomyData, `[[`, "superkingdom"),
    phagcn_phylum = sapply(TaxonomyData, `[[`, "phylum"),
    phagcn_class = sapply(TaxonomyData, `[[`, "class"),
    phagcn_order = sapply(TaxonomyData, `[[`, "order"),
    phagcn_family = sapply(TaxonomyData, `[[`, "family"),
    phagcn_genus = sapply(TaxonomyData, `[[`, "genus"),
    phagcn_species = sapply(TaxonomyData, `[[`, "species")
  ) %>%
  select(Accession, phagcn_superkingdom, phagcn_phylum, phagcn_class, phagcn_order, phagcn_family, phagcn_genus) 

colnames(final_phagcn)[1] <- "contigID"
```

CrassUS
```{r}
crassus <- vroom("~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/crassus_results.tsv")

final_crassus <- crassus %>%
  filter(discard == "FALSE") %>%
  filter(family != "outgroup") %>%
  mutate(
    crassus_superkingdom = "Viruses",
    crassus_phylum = "Uroviricota",
    crassus_class = "Caudoviricetes",
    crassus_order = "Crassvirales"
  ) %>%
  rename(
    contigID = contig,
    crassus_family = family,
    crassus_genus = genus,
    crassus_species = species
  ) %>%
  select(
    contigID, crassus_superkingdom, crassus_phylum, 
    crassus_class, crassus_order, crassus_family, crassus_genus, crassus_species
  ) %>%
  mutate(across(everything(), ~ifelse(. == "unknown", NA, .)))
```

Taxonomic merging and consensus 
```{r}
colnames(final_vc3)
colnames(final_mmseqs)
colnames(final_phagcn)
colnames(final_crassus)

# Merge all datasets by contigID
merged_taxonomies <- final_crassus %>%
  full_join(final_mmseqs, by = "contigID") %>%
  full_join(final_phagcn, by = "contigID") %>%
  full_join(final_vc3, by = "contigID") 

consensus_taxonomy <- merged_taxonomies %>%
  # Step 1: Determine the consensus superkingdom
  mutate(superkingdom = coalesce(crassus_superkingdom, 
                                 blast_superkingdom, 
                                 phagcn_superkingdom, 
                                 vc3_superkingdom)) %>%
  mutate(superkingdom = ifelse(is.na(superkingdom), "Viruses", superkingdom)) %>%
  # Step 2: Define NA-aware hierarchical checks
  mutate(
    # PHYLUM
    phylum = coalesce(
      ifelse(
        (crassus_superkingdom == superkingdom) | (is.na(crassus_superkingdom) & is.na(superkingdom)),
        crassus_phylum, NA
      ),
      ifelse(
        (blast_superkingdom == superkingdom) | (is.na(blast_superkingdom) & is.na(superkingdom)),
        blast_phylum, NA
      ),
      ifelse(
        (phagcn_superkingdom == superkingdom) | (is.na(phagcn_superkingdom) & is.na(superkingdom)),
        phagcn_phylum, NA
      ),
      ifelse(
        (vc3_superkingdom == superkingdom) | (is.na(vc3_superkingdom) & is.na(superkingdom)),
        vc3_phylum, NA
      )
    ),
    # CLASS (FIXED: Added missing closing parentheses)
    class = coalesce(
      ifelse(
        ((crassus_superkingdom == superkingdom) | (is.na(crassus_superkingdom) & is.na(superkingdom))) &
        ((crassus_phylum == phylum) | (is.na(crassus_phylum) & is.na(phylum))),
        crassus_class, NA
      ),
      ifelse(
        ((blast_superkingdom == superkingdom) | (is.na(blast_superkingdom) & is.na(superkingdom))) &
        ((blast_phylum == phylum) | (is.na(blast_phylum) & is.na(phylum))),
        blast_class, NA
      ),
      ifelse(
        ((phagcn_superkingdom == superkingdom) | (is.na(phagcn_superkingdom) & is.na(superkingdom))) &
        ((phagcn_phylum == phylum) | (is.na(phagcn_phylum) & is.na(phylum))),
        phagcn_class, NA
      ),
      ifelse(
        ((vc3_superkingdom == superkingdom) | (is.na(vc3_superkingdom) & is.na(superkingdom))) &
        ((vc3_phylum == phylum) | (is.na(vc3_phylum) & is.na(phylum))),
        vc3_class, NA
      )
    ),
    # ORDER
    order = coalesce(
      ifelse(
        ((crassus_superkingdom == superkingdom) | (is.na(crassus_superkingdom) & is.na(superkingdom))) &
        ((crassus_phylum == phylum) | (is.na(crassus_phylum) & is.na(phylum))) &
        ((crassus_class == class) | (is.na(crassus_class) & is.na(class))),
        crassus_order, NA
      ),
      ifelse(
        ((blast_superkingdom == superkingdom) | (is.na(blast_superkingdom) & is.na(superkingdom))) &
        ((blast_phylum == phylum) | (is.na(blast_phylum) & is.na(phylum))) &
        ((blast_class == class) | (is.na(blast_class) & is.na(class))),
        blast_order, NA
      ),
      ifelse(
        ((phagcn_superkingdom == superkingdom) | (is.na(phagcn_superkingdom) & is.na(superkingdom))) &
        ((phagcn_phylum == phylum) | (is.na(phagcn_phylum) & is.na(phylum))) &
        ((phagcn_class == class) | (is.na(phagcn_class) & is.na(class))),
        phagcn_order, NA
      ),
      ifelse(
        ((vc3_superkingdom == superkingdom) | (is.na(vc3_superkingdom) & is.na(superkingdom))) &
        ((vc3_phylum == phylum) | (is.na(vc3_phylum) & is.na(phylum))) &
        ((vc3_class == class) | (is.na(vc3_class) & is.na(class))),
        vc3_order, NA
      )
    ),
    # FAMILY
    family = coalesce(
      ifelse(
        ((crassus_superkingdom == superkingdom) | (is.na(crassus_superkingdom) & is.na(superkingdom))) &
        ((crassus_phylum == phylum) | (is.na(crassus_phylum) & is.na(phylum))) &
        ((crassus_class == class) | (is.na(crassus_class) & is.na(class))) &
        ((crassus_order == order) | (is.na(crassus_order) & is.na(order))),
        crassus_family, NA
      ),
      ifelse(
        ((blast_superkingdom == superkingdom) | (is.na(blast_superkingdom) & is.na(superkingdom))) &
        ((blast_phylum == phylum) | (is.na(blast_phylum) & is.na(phylum))) &
        ((blast_class == class) | (is.na(blast_class) & is.na(class))) &
        ((blast_order == order) | (is.na(blast_order) & is.na(order))),
        blast_family, NA
      ),
      ifelse(
        ((phagcn_superkingdom == superkingdom) | (is.na(phagcn_superkingdom) & is.na(superkingdom))) &
        ((phagcn_phylum == phylum) | (is.na(phagcn_phylum) & is.na(phylum))) &
        ((phagcn_class == class) | (is.na(phagcn_class) & is.na(class))) &
        ((phagcn_order == order) | (is.na(phagcn_order) & is.na(order))),
        phagcn_family, NA
      ),
      ifelse(
        ((vc3_superkingdom == superkingdom) | (is.na(vc3_superkingdom) & is.na(superkingdom))) &
        ((vc3_phylum == phylum) | (is.na(vc3_phylum) & is.na(phylum))) &
        ((vc3_class == class) | (is.na(vc3_class) & is.na(class))) &
        ((vc3_order == order) | (is.na(vc3_order) & is.na(order))),
        vc3_family, NA
      )
    ),
    # GENUS
    genus = coalesce(
      ifelse(
        ((crassus_superkingdom == superkingdom) | (is.na(crassus_superkingdom) & is.na(superkingdom))) &
        ((crassus_phylum == phylum) | (is.na(crassus_phylum) & is.na(phylum))) &
        ((crassus_class == class) | (is.na(crassus_class) & is.na(class))) &
        ((crassus_order == order) | (is.na(crassus_order) & is.na(order))) &
        ((crassus_family == family) | (is.na(crassus_family) & is.na(family))),
        crassus_genus, NA
      ),
      ifelse(
        ((blast_superkingdom == superkingdom) | (is.na(blast_superkingdom) & is.na(superkingdom))) &
        ((blast_phylum == phylum) | (is.na(blast_phylum) & is.na(phylum))) &
        ((blast_class == class) | (is.na(blast_class) & is.na(class))) &
        ((blast_order == order) | (is.na(blast_order) & is.na(order))) &
        ((blast_family == family) | (is.na(blast_family) & is.na(family))),
        blast_genus, NA
      ),
      ifelse(
        ((phagcn_superkingdom == superkingdom) | (is.na(phagcn_superkingdom) & is.na(superkingdom))) &
        ((phagcn_phylum == phylum) | (is.na(phagcn_phylum) & is.na(phylum))) &
        ((phagcn_class == class) | (is.na(phagcn_class) & is.na(class))) &
        ((phagcn_order == order) | (is.na(phagcn_order) & is.na(order))) &
        ((phagcn_family == family) | (is.na(phagcn_family) & is.na(family))),
        phagcn_genus, NA
      ),
      ifelse(
        ((vc3_superkingdom == superkingdom) | (is.na(vc3_superkingdom) & is.na(superkingdom))) &
        ((vc3_phylum == phylum) | (is.na(vc3_phylum) & is.na(phylum))) &
        ((vc3_class == class) | (is.na(vc3_class) & is.na(class))) &
        ((vc3_order == order) | (is.na(vc3_order) & is.na(order))) &
        ((vc3_family == family) | (is.na(vc3_family) & is.na(family))),
        vc3_genus, NA
      )
    ),
    # SPECIES (only crassus and blast)
    species = coalesce(
      ifelse(
        ((crassus_superkingdom == superkingdom) | (is.na(crassus_superkingdom) & is.na(superkingdom))) &
        ((crassus_phylum == phylum) | (is.na(crassus_phylum) & is.na(phylum))) &
        ((crassus_class == class) | (is.na(crassus_class) & is.na(class))) &
        ((crassus_order == order) | (is.na(crassus_order) & is.na(order))) &
        ((crassus_family == family) | (is.na(crassus_family) & is.na(family))) &
        ((crassus_genus == genus) | (is.na(crassus_genus) & is.na(genus))),
        crassus_species, NA
      ),
      ifelse(
        ((blast_superkingdom == superkingdom) | (is.na(blast_superkingdom) & is.na(superkingdom))) &
        ((blast_phylum == phylum) | (is.na(blast_phylum) & is.na(phylum))) &
        ((blast_class == class) | (is.na(blast_class) & is.na(class))) &
        ((blast_order == order) | (is.na(blast_order) & is.na(order))) &
        ((blast_family == family) | (is.na(blast_family) & is.na(family))) &
        ((blast_genus == genus) | (is.na(blast_genus) & is.na(genus))),
        blast_species, NA
      )
    )
  ) %>%
  # Final selection of columns
  select(contigID, superkingdom, phylum, class, order, family, genus, species)
```

Counts per taxonomic level after generating the consensus
```{r}
 tax_levels <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")

get_tax_counts <- function(data, tax_level) {
  data %>%
    group_by(!!sym(tax_level)) %>%
    count(!!sym(tax_level)) %>%
    ungroup() %>%
    rename_with(.cols = everything(), .fn = ~ paste(tax_level, ., sep = "_"))
}

list_of_counts <- lapply(tax_levels, function(tax) get_tax_counts(consensus_taxonomy, tax))

all_tax_counts <- bind_rows(list_of_counts, .id = "tax_level")
```

## Lifestyle Prediction: 

PHACTS
```{r}
phacts <- vroom(file = "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/PHACTS_concatenated_output.txt", col_names = FALSE)
phacts <- phacts %>%
  separate(X1, into = c("contigID", "phacts_lifestyle"), sep = " ") %>%
  rename(probability = X2, stdv = X3)
```

PhaTyp
```{r}
phatyp <- vroom::vroom("~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/phatyp_prediction.tsv", col_names = TRUE)
phatyp <- phatyp %>%
  select(-Length) 
colnames(phatyp) <- c("contigID", "phatyp_lifestylePred", "phatyp_lifestyleScore")
phatyp <- phatyp %>%
  mutate(phatyp_lifestylePred = ifelse(phatyp_lifestylePred == "-", NA, phatyp_lifestylePred),
         phatyp_lifestylePred = ifelse(phatyp_lifestylePred == "virulent", "Lytic", phatyp_lifestylePred))
```

Merging Lifestyle Predictions
```{r}
final_lifestyle <- phacts %>% 
  left_join(phatyp, by = "contigID")

consensus_lifestyle <- final_lifestyle %>%
  mutate(
    consensus_lifestyle = if_else(
      !is.na(phatyp_lifestylePred),
      phatyp_lifestylePred, 
      phacts_lifestyle
    ),
    agreement = case_when(
      phatyp_lifestylePred == phacts_lifestyle ~ "Match",
      is.na(phatyp_lifestylePred) | phatyp_lifestylePred == "" ~ "Only PHACTS",
      TRUE ~ "Discordant"
    )
  ) %>%
  select(
    contigID,
    phacts_lifestyle,
    phatyp_lifestylePred,
    consensus_lifestyle,
    agreement
  ) %>%
  mutate(consensus_lifestyle = ifelse(consensus_lifestyle == "temperate", "Temperate", consensus_lifestyle))

consensus_lifestyle %>% group_by(consensus_lifestyle) %>% count(agreement)
consensus_lifestyle <- consensus_lifestyle %>%
  select(contigID, consensus_lifestyle) 
```



## Host Prediction:iPHoP
```{r}
iphop <- vroom(file = "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/iphopPred_allCombined_toGenus_final.csv")

colnames(iphop) <- c("contigID", "AAI_to_closest_RaFAH_reference", "Host_lineage", "Confidence_score", "List_of_methods")

iphop <- iphop %>%
  separate(col = Host_lineage,
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus"),
           sep = ";",
           remove = FALSE) %>%
  mutate(across(.cols = c(Domain, Phylum, Class, Order, Family, Genus),
                .fns = ~ sub("^[a-z]__", "", .))) %>%
  mutate(FamGen = paste0(Family, "_", Genus)) %>%
  select(contigID, Phylum, Family, Genus) %>%
  filter(contigID != "Virus")

colnames(iphop)[1] <- "contigID"
```


## Abundance Table Construction

Loading coverage and abundace lon-format file
```{r}
abd_cov <- vroom("~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/08_Coverm_concatenated.txt", col_names = F)
# 
colnames(abd_cov) <- c("sampleID", "contigID", "rpkm", "tpm", "count", "variance", "mean", "covered_fraction", "covered_bases")
```

Abundance table using TPM 
```{r}
abd_cov_filt <- abd_cov %>%
  mutate(newTPM = ifelse(covered_fraction >= 0.75, tpm, 0)) %>%
  mutate(sampleID = gsub("_stats", "", sampleID))

newtpm_finalTable_phages <- reshape2::dcast(abd_cov_filt,contigID ~ sampleID,value.var =  "newTPM")
 
#Removing the contigs with 0 total abundance after changing abd values according to coverage 
newtpm_finalTable_phages <- newtpm_finalTable_phages %>%
  mutate(suma = rowSums(across(where(is.numeric)))) %>%
  filter(suma > 0) %>%
  select(-suma)
```

Abundance table using reads 
```{r}
abd_cov_filt <- abd_cov %>%
  mutate(newCount = ifelse(covered_fraction >= 0.75, count, 0)) %>%
  mutate(sampleID = gsub("_stats", "", sampleID))

newCount_finalTable_phages <- reshape2::dcast(abd_cov_filt,contigID ~ sampleID,value.var =  "newCount")
 
#Removing the contigs with 0 total abundance after changing abd values according to coverage 
newCount_finalTable_phages <- newCount_finalTable_phages %>%
  mutate(suma = rowSums(across(where(is.numeric)))) %>%
  filter(suma > 0) %>%
  select(-suma)
```


## pso building

Prepare metadata 
```{r}
# metadata_old <- readxl::read_xlsx(path = "/Users/luischica/Handley\ Lab\ Dropbox/Manuscripts/RC2/metadata/", na = "NA")

metadata <- vroom("/Users/luischica/Handley\ Lab\ Dropbox/Manuscripts/RC2/metadata/21032025_Metadata_freeze4_imputed_fin.txt")

META <- column_to_rownames(metadata, var = "database_ID")
META <- sample_data(META)
```

Using TPM
```{r}
#Convert dbl to interger and round counts (this helps with later steps)
phylo_countTable <- newtpm_finalTable_phages %>%
  column_to_rownames(var = "contigID") %>%
  mutate(across(
    .fns = ~ round(as.integer(.x))))
#Transpose and convert to matrix
phylo_countTable <- as.matrix(phylo_countTable)
class(phylo_countTable)

#Prepare TAX table
phylo_Final_Taxonomy <- consensus_taxonomy %>%
  column_to_rownames(var = "contigID")

phylo_Final_Taxonomy <- as.matrix(phylo_Final_Taxonomy)
class(phylo_Final_Taxonomy)

#Create Phyloseq Object
library(phyloseq)
OTU <- otu_table(phylo_countTable, taxa_are_rows = TRUE)
TAX <- tax_table(phylo_Final_Taxonomy)

ps0 <- phyloseq(OTU, TAX)
ps0

#Final Phyloseq object 
VirphyloseqOBJ <- merge_phyloseq(ps0,META)

library(microViz)
VirphyloseqOBJ <- VirphyloseqOBJ %>% tax_fix()

VirphyloseqOBJ

saveRDS(VirphyloseqOBJ, "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/outputData/VirphyloseqOBJ.rds")
```

pso using counts and not tpm. For diff abd testing 
```{r}
#Convert dbl to interger and round counts (this helps with later steps)
phylo_countTable <- newCount_finalTable_phages %>%
  column_to_rownames(var = "contigID") %>%
  mutate(across(
    .fns = ~ round(as.integer(.x))))
#Transpose and convert to matrix
phylo_countTable <- as.matrix(phylo_countTable)
class(phylo_countTable)

#Prepare TAX table
phylo_Final_Taxonomy <- consensus_taxonomy %>%
  column_to_rownames(var = "contigID")

phylo_Final_Taxonomy <- as.matrix(phylo_Final_Taxonomy)
class(phylo_Final_Taxonomy)

#Create Phyloseq Object
OTU <- otu_table(phylo_countTable, taxa_are_rows = TRUE)
TAX <- tax_table(phylo_Final_Taxonomy)

ps0 <- phyloseq(OTU, TAX)
ps0

#Final Phyloseq object 
VirphyloseqOBJ_counts <- merge_phyloseq(ps0,META)

VirphyloseqOBJ_counts <- VirphyloseqOBJ_counts %>% tax_fix()

VirphyloseqOBJ_counts

saveRDS(VirphyloseqOBJ_counts, "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/outputData/VirphyloseqOBJ_counts.rds")
```

checking sample discrepancies between final pso and metadata. 
```{r}
meta_pso <- data.frame(sample_data(VirphyloseqOBJ)) %>%
  rownames_to_column(var = "database_ID")

metaMissing <- metadata %>%
  filter(database_ID %notin% meta_pso$database_ID)
```
All samples contained in metadata file and not present in the pso file have no virome runs, except for 35611, lost in coverage filter. 

Final NOTE: After running all the chunks, we should get 3 output files: 
  VirphyloseqOBJ_counts.rds
  VirphyloseqOBJ.rds
  contigInfo.txt


## Generate contig Info Table
```{r}
checkV <-  vroom::vroom("~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/inputData/PhageContigs_quality_summary.tsv", delim = "\t", col_names = T)
colnames(checkV)[1] <- "contigID"
checkV <- checkV %>%
  filter(contig_length >= 1000)

contigInfo <- checkV %>%
  tidylog::left_join(consensus_lifestyle, by = "contigID") %>%
  tidylog::left_join(iphop, by = "contigID") %>%
  tidylog::left_join(consensus_taxonomy, by = "contigID") %>%
  dplyr::distinct(contigID, .keep_all = T)

contigInfo <- contigInfo %>%
  filter(contigID %in% newtpm_finalTable_phages$contigID) %>%
  distinct(contigID, .keep_all = TRUE)

vroom::vroom_write(contigInfo, "~/Handley\ Lab\ Dropbox/Manuscripts/RC2/phage/01_psoConstructoin_genomicInfoFormating/outputData/contigInfo.txt")
```
